---
title: "NYPD Shootings: Murder Rates of Single- and Multi-victim Shootings"
author: "Ron Sielinski"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  word_document: default
  html_document: default
data_URL: "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

New York City has kept track of all the shootings that have taken place within its boroughs since 2006. This analysis compares the murder rates of single- and multi-victim shootings.

## Motivating Hypothesis

The more people who get shot in an incident, the more likely that someone will die. That makes sense mathematically: If only one person gets shot, only one person has a chance of dying. But if twelve people get shot, twelve people have a chance of dying. 

But that doesn't necessarily mean that multi-victim incidents are more deadly on a *per-victim* basis. If a shooting involves only one victim, there's a good chance that the perpetrator will have meant to harm the victim, increasingly the likelihood of fatal injuries. In a multi-victim shooting, however, there's a reasonable chance that some of the victims will have been innocent bystanders and their injuries incidental. On a per-victim basis, we might actually find that single-victim shootings are *more* deadly. 

This analysis will explore over 20,000 shootings that taken place in New York City since 2006 to find out. 

## Raw Data

The City of New York publishes its shooting data at <https://data.cityofnewyork.us>.

```{r load_data, warning = F, message = F}

###################
## Load Packages
###################

library(tidyverse)

###################
## Load the Data
###################

# load data from the City of New York
data_file <- 'https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD'
ny_dat <- read_csv(data_file)

# clean up the environment
rm(data_file)

```

## Preparing the Data

The city also publishes a guide to the data at <https://data.cityofnewyork.us/Public-Safety/NYPD-Shooting-Incident-Data-Historic-/833y-fsy8/about_data>. For this analysis, we only need a small subset of the data: 

* `INCIDENT_KEY`: A unique identifier for each incident
* `STATISTICAL_MURDER_FLAG`: A Boolean that indicates whether the shooting resulted in the victim’s death
* `PERP_AGE_GROUP`, `PERP_SEX`, `PERP_RACE`: A set of factors that describe the perpetrator 

```{r tidy_data}

###################
## Tidy the Data
###################

# for convenience, convert column names to lower case (to make typing easier)
colnames(ny_dat) <- colnames(ny_dat) |> 
  str_to_lower()

# perpetrator factors
ny_dat$perp_age_group <- as.factor(ny_dat$perp_age_group)
ny_dat$perp_sex <- as.factor(ny_dat$perp_sex)
ny_dat$perp_race <- as.factor(ny_dat$perp_race)

# keep only fields needed to analyze multi-victim incidents
ny_dat <- ny_dat |> 
  select(incident_key, statistical_murder_flag, starts_with('perp'))

```

## Exploring the Data

### Victims

The first question that we want to answer is how many incidents have more than one victim?

```{r EDA}

#########
## EDA
#########

# create a data frame that summarizes the number of victims and murders for 
# each incident
incident_cnts <- ny_dat |>
  summarize(victims = n(), 
            murders = sum(statistical_murder_flag), 
            .by = incident_key) |> 
  arrange(desc(victims), desc(murders))

head(incident_cnts)

```
The results show that one incident had eighteen victims!

However, there is only one eighteen-victim incident out of `r nrow(incident_cnts)` incidents, which raises questions about the typical number of victims and murders involved in each incident. 

```{r victims_dist}

# look at the frequency distribution of victims per incident 
incident_cnts |> 
  count(victims) |>
  ggplot(aes(x = victims, y = n)) +
  geom_bar(stat = 'identity') +  
  scale_x_continuous(breaks = pretty(incident_cnts$victims, 
                                     n = length(unique(incident_cnts$victims)))) +
  geom_text(aes(x = victims, y = n, label = n), vjust = -0.5) +
  labs(title = 'Frequency Distribution of Victims per Incident', 
       y = 'Count of Incidents', 
       x = 'Victims per Incident')

```

Plotting the frequency distribution of the victims per incident reveals that most incidents have only one victim. (By definition, an incident involves at least one victim, so no incidents can have zero victims.) 

Given the nature of the data (i.e., a count of victims) and the shape of the curve, the count of victims per incident appears to follow a Poisson distribution. 

### Murders

```{r murders_dist}

# look at the frequency distribution of murders per incident 
# we'll reuse this plot, so store it into the environment
murders_plot <- incident_cnts |>
  count(murders) |>
  ggplot(aes(x = murders, y = n)) +
  geom_bar(stat = 'identity') +
  scale_x_continuous(breaks = pretty(incident_cnts$murders, 
                                     n = length(unique(incident_cnts$murders)))) +
  geom_text(aes(x = murders, y = n, label = n), vjust = -0.5) +
  labs(title = 'Frequency Distribution of Murders per Incident', 
       y = 'Count of Incidents', 
       x = 'Murders per Incident')

murders_plot

```

A frequency distribution plot of the murders per incident looks again like a Poisson distribution: The count of incidents decreases rapidly as the murders per incident increases.  

In this case, however, it's possible for incidents to have no murders, which is most often the case. 

A final question to explore is the number of perpetrators (or shooters) per incident. The data doesn't specifically count or uniquely identify perpetrators, but for each incident, we can use any changes in the perpetrator(s) age, sex, or race to determine whether multiple shooters were involved. 

### Shooters

Here's a single incident that had shooters from five separate demographic groups.

```{r multi_shooter_example, echo = F}

# look at an example of an incident with multiple shooters
ny_dat |>
  filter(incident_key == 90128858) |>
  select(incident_key, perp_age_group, perp_sex, perp_race)

```

The obvious limitation with this approach is that it's unable to distinguish between perpetrators who are the same age, sex, and race. As such, the number of differences that we detect across the set of factors will be the *minimum* number of shooters.   

```{r perps}

# for each incident, count the number of victims with the same set of perp factors
# each row in ny_dat is a victim, so the count of rows is the count of victims
perp_factor_cnts <- ny_dat |> 
  summarize(victims_by_perp_factors = n(), 
            murders_by_perp_factors = sum(statistical_murder_flag), 
            .by = c(incident_key, perp_age_group, perp_sex, perp_race))

# for each incident, count the number of perp factor sets and the total number 
# of victims and murders
factor_cnts  <- perp_factor_cnts |>
  summarize(factor_sets = n(), 
            victims_cnt = sum(victims_by_perp_factors), 
            murders_cnt = sum(murders_by_perp_factors), 
            .by = incident_key) 

# clean up the environment
rm(perp_factor_cnts)

# look at incidents that have involved the most number of shooters
factor_cnts |>
  rename(shooter_cnt_min = factor_sets) |>
  arrange(desc(shooter_cnt_min)) |>
  head()

```

The result shows us that multiple incidents have involved five (or more) shooters, including incidents where  of the shooters may have all murdered each other. 

This adds a new dimension to the original hypothesis. Without explicitly stating it, we had presumed that all incidents were perpetrated by single shooters. However, given that there were incidents that involved multiple shooters, there's a good chance that multiple guns were fired---potentially at close range---increasing the chance of death for everyone involved. 

## Statistical Models

### Per-victim Model

Because the above show plots focus on just victims and just murders, they don't offer much insight into the  relationship is between victims and murders. 

Plotting the average number of murders by the number of victims shows us that, as the number of victims (*n*) increases, so does the average number of murders. 

It's worth noting, that for some values of *n*, there were no murders. There were only two seven-victim shootings and two nine-victim shootings. That draws attention to the fact that for large value of *n*, we might not have enough data to draw statistically significant inferences.  

```{r actuals_bar}

#########################
## Bar Plot of Actuals
#########################

# range for the number of victims (n)
n_victims <- min(incident_cnts$victims):max(incident_cnts$victims)
# range for the number of murders
n_murders <- min(incident_cnts$murders):max(incident_cnts$murders)

# calculate the average number of murders for each n
murder_rates <- incident_cnts |>
  summarize(incidents = n(),
            murders = sum(murders),
            # average number of murders  
            per_incident_rate = sum(murders) / n(),
            .by = victims) |>
  # observed murders per victim
  mutate(per_victim_rate = per_incident_rate / victims) |>
  arrange(victims)

# plot the result
murder_rates |>
  select(victims, starts_with('per_incident')) |>
  pivot_longer(cols = starts_with('per_incident'),
               names_to = 'rate_type',
               values_to = 'rate') |>
  ggplot(aes(x = factor(victims), y = rate, fill = rate_type)) +
  geom_bar(stat = 'identity', position = position_dodge(width = 0.9)) +
  #geom_text(aes(x = factor(victims), y = rate, label = sprintf("%.2f", rate)), 
  #          position = position_dodge(width = 0.9), 
  #          hjust = 1.25,
  #          angle = 90, 
  #          color = 'white') +
  scale_fill_manual(values = c('black'), labels = c('Observed Rate')) +
  labs(title = 'Average Number of Murders by Number of Victims',
       y = 'Average Number of Murders',
       x = 'Number of Victims (n)')

```

Of course, the number of murders is always capped by the number of victims, so the average for an *n*-victim murder can never exceed *n*. To adjust for the number of victims, we can calculate a global murders-to-victims rate.  

We can then multiply the per-victim rate by *n* to model the number of murders we can expect for each *n*. 
However, a plot of the observed and expected values suggests that the resulting model performs poorly, overestimating the number of murders for small values of *n* and underestimating the number of murders for large *n*.   

```{r global_rate}

####################################
## Global Murders-to-victims Rate
####################################

# calculate a global murders-to-victims rate
global_murders_per_victim <- sum(incident_cnts$murders) / sum(incident_cnts$victims)

# use global_rate to predict the number of murders for each n
murder_rates <- murder_rates |>
  mutate(
    per_incident_global = victims * global_murders_per_victim, 
    per_victim_global = global_murders_per_victim, 
    murders_global = incidents * per_incident_global
  )

# plot the results
murder_rates |>
  select(victims, starts_with('per_incident')) |>
  pivot_longer(cols = starts_with('per_incident'),
               names_to = 'rate_type',
               values_to = 'rate') |>
  ggplot(aes(x = factor(victims), y = rate, fill = rate_type)) +
  geom_bar(stat = 'identity', position = position_dodge(width = 0.9)) +
  #geom_text(aes(x = factor(victims), y = rate, label = sprintf("%.2f", rate)), 
  #          position = position_dodge(width = 0.9), 
  #          hjust = 1.25,
  #          angle = 90, 
  #          color = 'white') +
  scale_fill_manual(name = 'Rate/Model', 
                    values = c('darkgrey', 'black'), labels = c('Per-victim Rate ', 'Observed Rate')) +
  labs(title = 'Average Number of Murders by Number of Victims',
       y = 'Average Number of Murders',
       x = 'Number of Victims (n)')

```

To quantify the model's performance, we can calculate its Root Mean Squared Error (RMSE). 

```{r RMSE}

# Calculate the root mean squared error
RMSE <- function(actual_values, estimated_values) {
  ((actual_values - estimated_values) ^ 2) |>
    mean() |>
    sqrt()
}

RMSE(incident_cnts$murders, incident_cnts$victims * global_murders_per_victim)

```
The RMSE for estimates based on the per-victim model is `r round(RMSE(incident_cnts$murders, incident_cnts$victims * global_murders_per_victim), 2)`. 

### Per-incident Model

Alternatively, we can model the number of murders using a Poisson distribution. In doing so, we shift our perspective slightly from the murders-to-victims rate, which is the average number of murders per *victim*, to the average number of murders per *incident*. 

We can then use the properties of a Poisson distribution to model the number of murders that we can expect for each *n*. 

```{r pois_model}

################################################
## model murders using a Poisson distribution
################################################

cnt_all_incidents <- nrow(incident_cnts)

# average number of murders per incident 
murders_lambda <- mean(incident_cnts$murders)

# we'll want to compare the model results with actuals, so count the incidents 
# and sum the murders for each number of murders (n)
pois_model <- incident_cnts |> 
  summarize(incidents_cnt = n(), 
            murders_cnt = sum(murders), 
            .by = murders) |> 
  # calculate the percentage of incidents that had n murders
  mutate(incidents_pct = incidents_cnt / cnt_all_incidents) 

# add rows to the data frame for values of n that had no murders 
missing_n <- data.frame(murders = n_murders[!(n_murders %in% pois_model$murders)], 
                        incidents_cnt  = 0, 
                        murders_cnt = 0, 
                        incidents_pct  = 0
)
pois_model <- pois_model |>
  bind_rows(missing_n) |> 
  arrange(murders)

# use the Poisson model to predict the percentage of incidents with n murders
# and the resulting number of incidents 
pois_model <- pois_model |>
  mutate(model_pct = dpois(murders, lambda = murders_lambda), 
         incidents_pred = round(cnt_all_incidents * model_pct, 0), 
         murders_pred = incidents_pred * murders)
 
# confirm that the model accounts for all incidents (numbers should be close)
#sum(pois_model$incidents_cnt)
#sum(pois_model$incidents_pred)

# confirm that the model accounts for all murders (numbers should be close)
#sum(pois_model$murders_cnt)
#sum(pois_model$murders_pred)

# clean up the environment
rm(missing_n)

# plot the data and the model
murders_plot +
  geom_point(data = pois_model, aes(x = murders, y = incidents_pred), color = 'red') +
  geom_line(data = pois_model, aes(x = murders, y = incidents_pred), color = 'red')

```
### Validate the Poisson Model

We can test the model's goodness of fit with a chi-squared test, which compares the observed and expected percentage of murders for each *n*. 

```{r pois_validate, warning = F}

####################
## Validate model
####################

# perform the chi-squared test
# the expected counts/frequencies can't be too small (ideally, >= 5)
# one workaround is to combine bins
max_expected_bin <- max(which(pois_model$incidents_pred >= 5))

# for observed incidents, get the percentages for each bin up to the shared_max 
# and combine whatever remains into another bin
incidents_pcts <- c(pois_model$incidents_pct[1:max_expected_bin], 
                    sum(pois_model$incidents_pct[(max_expected_bin + 1):nrow(pois_model)])
)

# do the same for the modeled bins
model_pcts <- c(pois_model$model_pct[1:max_expected_bin], 
                sum(pois_model$model_pct[(max_expected_bin + 1):nrow(pois_model)])
)

# A high p-value suggests that the observed data is consistent with the Poisson 
# distribution, within the bounds of normal statistical variation
chi_squared_test <- chisq.test(x = incidents_pcts, p = model_pcts)

# remember, we're looking at the first few bins (and everything else is binned together)
print(chi_squared_test)

```

A high p-value suggests that the observed data is consistent with the expected values generated by our Poisson model (within the bounds of normal statistical variation). In this case, the p-value is `r round(chi_squared_test$p.value, 3)`, which is reassuringly high. 

Note, however, that we could only test the model's first `r max_expected_bin` bins, because the number of incidents with >`r (max_expected_bin) - 1` murders was too small to include in the test. (All incidents with >`r (max_expected_bin) - 1` murders were combined into a single bin.) Nonetheless, incidents with <`r (max_expected_bin)` murders comprise `r round(sum(pois_model$incidents_pct[0:max_expected_bin]) * 100, 1)`% of incidents, so incidents with >`r (max_expected_bin) - 1` murders are low-probability outliers. 

```{r pois_plot}

# add results from the Poisson model to the results data frame
murder_rates <- murder_rates |>
  mutate(
    # use the Poisson model to predict the number of murders
    per_incident_pois = victims * murders_lambda, 
    per_victim_pois = per_incident_pois / victims,
    murders_pois = incidents * per_incident_pois 
  )

# plot the results
murder_rates |>
  select(victims, starts_with('per_incident')) |>
  pivot_longer(cols = starts_with('per_incident'),
               names_to = 'rate_type',
               values_to = 'rate') |>
  ggplot(aes(x = factor(victims), y = rate, fill = rate_type)) +
  geom_bar(stat = 'identity', position = position_dodge(width = 0.9)) +
  #geom_text(aes(x = factor(victims), y = rate, label = sprintf("%.2f", rate)), 
  #          position = position_dodge(width = 0.9), 
  #          hjust = 1.25,
  #          angle = 90, 
  #          color = 'white') +
  scale_fill_manual(name = 'Rate/Model',
                    values = c('darkgrey', 'blue', 'black'), 
                    labels = c('Per-victim Rate', 'Per-incident Rate', 'observed Rate')) +
  labs(title = 'Average Number of Murders by Number of Victims',
       y = 'Average Number of Murders',
       x = 'Number of Victims (n)')

```

For *n* > 3, the per-incident model (i.e., Poisson model) comes closer to predicting the observed murder rates than the per-victim model. For small *n*, the per-victim model is better. 

Here, it's important not to confuse the *n* (the number of victims) with the bins in the chi-squared test (murders per incident). Again, incidents with <`r (max_expected_bin)` murders comprise `r round(sum(pois_model$incidents_pct[0:max_expected_bin]) * 100, 1)`% of incidents and could have been spread across all values of *n*.  

A more robust way to compare the overall performance of the per-incident and per-victim models is through their RSME values. 

Although the Poisson models uses the same murders-per-incident rate for every incident, `r round(murders_lambda, 2)` murders per incident, we can still calculate the model's RMSE. 

```{r pois_perf}

RMSE(incident_cnts$murders, rep(murders_lambda, nrow(incident_cnts)))

```

The per-incident model's RMSE is `r round(RMSE(incident_cnts$murders, rep(murders_lambda, nrow(incident_cnts))), 2)`, not quite as good as the per-victim model, which has an RMSE of `r round(RMSE(incident_cnts$murders, incident_cnts$victims * global_murders_per_victim), 2)`. 

### Linear Model

A linear model takes yet another approach, our final. The least squares method tries to capture the linear relationship between the number of victims and the number of murders across all data points, assuming each additional victim is associated with a *constant* change in the number of murders. 

```{r linear_model}

# Build a simple linear model, based solely on the number of victims
lm_fit <- lm(murders ~ victims, data = incident_cnts)
summary(lm_fit)

# use the model to predict the number of murders
murder_rates$per_incident_linear <- predict(lm_fit, newdata = murder_rates)

# calculate the per-victim rate
murder_rates <- murder_rates |>
  mutate(per_victim_linear = per_incident_linear / victims)

# plot the results
murder_rates |>
  select(victims, starts_with('per_incident')) |>
  pivot_longer(cols = starts_with('per_incident'),
               names_to = 'rate_type',
               values_to = 'rate') |>
  ggplot(aes(x = factor(victims), y = rate, fill = rate_type)) +
  geom_bar(stat = 'identity', position = position_dodge(width = 0.9)) +
  #geom_text(aes(x = factor(victims), y = rate, label = sprintf("%.2f", rate)), 
  #          position = position_dodge(width = 0.9), 
  #          hjust = 1.25,
  #          angle = 90, 
  #          color = 'white') +
  # add the linear model
  #geom_line(aes(y = linear_model), color = 'red', linetype = 'dashed') +  
  #geom_point(aes(y = linear_model), color = 'red') +  
  scale_fill_manual(name = 'Rate/Model', 
                    values = c('darkgrey', 'lightblue', 'blue', 'black'), 
                    labels = c('Per-victim Rate', 'Linear Model', 'Per-incident Rate', 'observed Rate')) +
  labs(title = 'Average Number of Murders by Number of Victims',
       y = 'Average Number of Murders',
       x = 'Number of Victims (n)')

```
```{r lm_perf}

RMSE(incident_cnts$murders, predict(lm_fit, newdata = incident_cnts))

```

The RMSE for the linear model is `r round(RMSE(incident_cnts$murders, predict(lm_fit, newdata = incident_cnts)), 2)`, better than both the per-incident model (`r round(RMSE(incident_cnts$murders, rep(murders_lambda, nrow(incident_cnts))), 2)`) and the per-victim model (`r round(RMSE(incident_cnts$murders, incident_cnts$victims * global_murders_per_victim), 2)`). 

## Conclusion

Previous charts showed the expected average for the number of murders for each *n*. The plot below normalizes the number of murders by the number of victims, effectively showing the expected per-victim murder rate for each *n*.    

```{r per-victim, warning = F}

#############################
## per-victim murder rates 
#############################

# plot the per-victim murder rates
murder_rates |>
  ggplot(aes(x = victims, y = per_victim_rate)) +
  geom_col(aes(fill = 'Observed Rate'), color = 'black') +
  # add the global per-victim rate
  geom_hline(aes(yintercept = global_murders_per_victim, color = 'Per-victim Model'), 
             linetype = 'twodash', linewidth = 1.2) +
  # add the Poisson model's rate
  geom_hline(aes(yintercept = murders_lambda, color = 'Per-incident (Poisson) Model'), 
             linetype = 'twodash', linewidth = 1.2) +
  # add the linear model's rate
  geom_line(aes(y = per_victim_linear, color = 'Linear Model'), 
            linetype = 'twodash', linewidth = 1.2) +
  scale_color_manual(
    name = 'Expected Rate',
    values = c('Per-victim Model' = 'darkgrey', 
               'Linear Model' = 'lightblue', 
               'Per-incident (Poisson) Model' = 'blue', 
               'Observed Rate' = 'black'),
    labels = c('Linear Model', 'Per-incident (Poisson) Model', 'Per-victim Model', 'Observed Rate')) +
  scale_fill_manual(
    name = 'Observed Rate',
    values = c('Observed Rate' = 'black'),
    labels = c('Observed Rate')) +
  labs(title = 'Per-victim Murder by the Number of victims', 
       subtitle = 'Observed and Expected Rates', 
       y = 'Per-victim Murder Rate', 
       x = 'Number of Victims (n)') 

```

The per-victim and per-incident models predict that the murder rate stays the same for all values of *n*, which would suggest that all shootings are equally deadly, regardless of *n*. The linear model, on the other hand, predicts that the murder rate increases as *n* increases, which would suggest that single-victim shootings are *less* deadly than multi-victim shootings. (The model predicts an increased murder rate for *every* increment of *n*.)

Based on RMSE alone, we should favor the linear model: However, the RMSE for the linear model (`r round(RMSE(incident_cnts$murders, predict(lm_fit, newdata = incident_cnts)), 2)`) is only slightly better than the per-victim model (`r round(RMSE(incident_cnts$murders, incident_cnts$victims * global_murders_per_victim), 2)`). 

Moreover, the observed rate in above plot suggests that both the per-victim and linear models have regions of localized strength. At a high-level, the observed rate is bifurcated---less than 0.2 murders per victim for *n* < 4, and ~0.4 murders per victim for *n* > 3---a difference of approximately 2x. The per-victim model appears to work best for *n* < 4, and the linear model appears to work best for for *n* > 3. 

It's worth noting, however, that the upward trend in the murder rate for larger values of *n* could be an artifact of small sample sizes and that the linear model overfit to the few available incidents. 

To mitigate the risk of small sample-bias elsewhere in this analysis, we used averages (i.e., global averages for the per-victim and per-incident models) and binning (i.e., combining bins with very few incidents to create a larger, more stable group for analysis). But nothing has been done to mitigate the risk of small-sample bias for the linear model. (The inclusion of confidence intervals might have made the effective of any small-sample bias more apparent.)

Other sources of bias:

* Most of the incidents (`r round(count(filter(incident_cnts, victims == 1)) / cnt_all_incidents * 100, 2)`%) involve only one victim, so single-victim incidents play a significant role in the overall RMSE of a model: Their errors will have a substantial cumulative impact on the RMSE. As the above plot illustrates, however, the expected murder rate of the per-victim model (`r round(murder_rates$per_victim_global[1], 2)`) is actually closer to the observed rate (`r round(murder_rates$per_victim_rate[1], 2)`) for single-victim incidents than the linear model (`r round(murder_rates$per_victim_linear[1], 2)`). 

* The online guide to the data specifically refers to `INCIDENT_KEY` as a "randomly generated persistent ID for each arrest". That suggests that an incident might be counted only if a perpetrator is arrested. That might make the analysis blind to shootings---and murders---where an arrest did *not* occur. 

* Likewise, the definition of "murder" is not clear. For example, does murder depend on a verdict or plea of guilty? Would the historic count of murders change if a perpetrator were acquitted? Similarly, is manslaughter (an *unintentional* killing) considered the same as---or different from---murder (an *intentional* killing)? A clear definition of "murder" is important for both a correct framing of the analysis and interpretation of results.

More research and analysis are required to address these issues, but the possibility exists that the best model for the murder rate is actually some combination of the three. Multiple use cases exist where the best "model" is actually a combination of models: mixture models, zero-inflated models, ensemble models, and more. 

Ultimately, the analysis in this paper suggests that single-victim shootings are *less* deadly than multi-victim incidents, but only when four or more victims are involved. When three or fewer victims are involved, single-victim shootings are approximately equally deadly. 
